---
title: "EV Power - Lab 4 Project Report"
format: typst
---

# Example Solution 1

## **Part 0: libraries**

```{r}
install.packages(c("dplyr", "stringr", "ggplot2", "sf", "leaflet"))

# Load the libraries
library(tidyverse)
library(sf)
library(leaflet)
```

## **Part 1:** **Defining Research Question**

Chosen Question:"How has the share of renewable energy changed from 2021–2023 across states?"

## **Part 2: Data Preparation and Cleaning**

```{r}

setwd("/Users/kruna/Desktop/college/year3/statistics133/project4/ev-power-krunamasic")

# Load the total energy use datasets for 2021, 2022, and 2023
total_use_2021 <- read.csv("data/total-use-2021.csv")
total_use_2022 <- read.csv("data/total-use-2022.csv")
total_use_2023 <- read.csv("data/total-use-2023.csv")

```

```{r}
#Clean Data


# Function to reshape and clean total energy data
clean_total_energy <- function(df, year) {
  # Pivot from wide to long format
  df_long <- df %>%
    # Pivot all columns except Energy_Source into long format
    pivot_longer(
      cols = -Energy_Source,           # All columns except Energy_Source
      names_to = "state",               # Column names become "state"
      values_to = "energy_value"        # Values go into "energy_value"
    ) %>%
    # Clean the Energy_Source names
    mutate(Energy_Source = str_trim(Energy_Source) %>% str_to_lower()) %>%
    # Add year column
    mutate(year = year)
  
  return(df_long)
}

# Apply to all three years
total_2021_long <- clean_total_energy(total_use_2021, 2021)
total_2022_long <- clean_total_energy(total_use_2022, 2022)
total_2023_long <- clean_total_energy(total_use_2023, 2023)

# Combine all years
total_all <- bind_rows(total_2021_long, total_2022_long, total_2023_long)

# Look at the result
print("Reshaped data sample:")
head(total_all, 20)
```


## **Part 3: Joining / Pivoting Datasets for Analysis**

### Step 1: Calculate total energy per state per year

```{r calculate-total-energy}
# Sum all energy sources to get total energy use
total_energy_by_state <- total_all %>%
  group_by(state, year) %>%
  summarize(total_energy = sum(energy_value, na.rm = TRUE), .groups = 'drop')

print("Total energy by state and year:")
head(total_energy_by_state, 10)
```

### Step 2: Extract renewable energy data

```{r extract-renewable}
## Filter to get only renewable energy
# Note: The name varies between years, so we use pattern matching
renewable_energy <- total_all %>%
  filter(str_detect(Energy_Source, "renewable")) %>%
  select(state, year, renewable_energy = energy_value)

print("Renewable energy by state and year:")
head(renewable_energy, 10)

# Check how many rows per year
print("Count by year:")
renewable_energy %>% count(year)
```

### Step 3: Join renewable with total energy and calculate percentage

```{r calculate-percentage}
# Join renewable energy with total energy
energy_combined <- renewable_energy %>%
  left_join(total_energy_by_state, by = c("state", "year"))

# Calculate renewable percentage
energy_combined <- energy_combined %>%
  mutate(renewable_pct = (renewable_energy / total_energy) * 100)

print("Combined data with renewable percentage:")
head(energy_combined, 10)

# Summary statistics
print("Summary of renewable percentages:")
summary(energy_combined$renewable_pct)
```


### Step 4: Calculate change from 2021 to 2023

```{r calculate-change}
# Get 2021 data
energy_2021 <- energy_combined %>%
  filter(year == 2021) %>%
  select(state, renewable_pct_2021 = renewable_pct)

# Get 2023 data
energy_2023 <- energy_combined %>%
  filter(year == 2023) %>%
  select(state, renewable_pct_2023 = renewable_pct)

# Calculate the change
energy_change <- energy_2021 %>%
  left_join(energy_2023, by = "state") %>%
  mutate(pct_point_change = renewable_pct_2023 - renewable_pct_2021)

print("Change in renewable percentage from 2021 to 2023:")
head(energy_change, 10)

# Summary of changes
print("Summary of percentage point changes:")
summary(energy_change$pct_point_change)

# Top 5 states with biggest increases
print("Top 5 states with largest increases:")
energy_change %>%
  arrange(desc(pct_point_change)) %>%
  head(5)

# Top 5 states with biggest decreases
print("Top 5 states with largest decreases:")
energy_change %>%
  arrange(pct_point_change) %>%
  head(5)
```

### Step 5: Prepare data for mapping

```{r prepare-for-mapping}
# Create a lookup table for state abbreviations to full names
state_lookup <- data.frame(
  state = state.abb,
  state_name = tolower(state.name)
)

# Add DC manually (not in state.abb)
state_lookup <- rbind(state_lookup, data.frame(state = "DC", state_name = "district of columbia"))

# Join with our data
energy_change_map <- energy_change %>%
  left_join(state_lookup, by = "state")

print("Data ready for mapping:")
head(energy_change_map, 10)

# Check for any states that didn't match
print("States without full names:")
energy_change_map %>%
  filter(is.na(state_name)) %>%
  select(state)
```

## **Part 4: Mapping Visualization**

### Get US map data

```{r get-map-data}
# Get US state map data
states_map <- map_data("state")

print("Map data sample:")
head(states_map)
```

### Join our energy data with map data

```{r join-map-data}
# Join our energy change data with the map coordinates
map_data_joined <- states_map %>%
  left_join(energy_change_map, by = c("region" = "state_name"))

print("Joined map data sample:")
head(map_data_joined)
```

### Create the map: Change in Renewable Energy Share (2021-2023)

```{r create-map, fig.width=12, fig.height=8}
# Create map showing change in renewable percentage
ggplot(map_data_joined, aes(x = long, y = lat, group = group, fill = pct_point_change)) +
  geom_polygon(color = "white", size = 0.3) +
  scale_fill_gradient2(
    low = "#d73027",           # Red for decreases
    mid = "#ffffbf",           # Yellow for no change
    high = "#1a9850",          # Green for increases
    midpoint = 0,
    name = "Change in\nRenewable %\n(percentage\npoints)",
    na.value = "grey80"
  ) +
  coord_fixed(1.3) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "grey40"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) +
  labs(
    title = "Change in Renewable Energy Share Across U.S. States (2021-2023)",
    subtitle = "How much did each state's renewable energy percentage change?"
  )
```

### Summary tables of biggest changes

```{r summary-table}
# Top 10 increases
print("States with Largest Increases in Renewable Energy Share:")
energy_change %>%
  arrange(desc(pct_point_change)) %>%
  head(10) %>%
  mutate(
    renewable_pct_2021 = round(renewable_pct_2021, 2),
    renewable_pct_2023 = round(renewable_pct_2023, 2),
    pct_point_change = round(pct_point_change, 2)
  ) %>%
  knitr::kable(
    col.names = c("State", "2021 (%)", "2023 (%)", "Change (pct pts)"),
    align = c("l", "r", "r", "r")
  )

# Top 10 decreases
print("\nStates with Largest Decreases in Renewable Energy Share:")
energy_change %>%
  arrange(pct_point_change) %>%
  head(10) %>%
  mutate(
    renewable_pct_2021 = round(renewable_pct_2021, 2),
    renewable_pct_2023 = round(renewable_pct_2023, 2),
    pct_point_change = round(pct_point_change, 2)
  ) %>%
  knitr::kable(
    col.names = c("State", "2021 (%)", "2023 (%)", "Change (pct pts)"),
    align = c("l", "r", "r", "r")
  )
```



